On Optimality of Finite Element Integration

### 2015-12-22 TOMS_Paper/assembly.tex
Improve cost function description

\subsection{Transformation algorithm}
We address the two following issues: 
\begin{enumerate}
\item \textit{Coordination of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation could either increase or decrease the operation count with respect to sharing elimination.
\item \textit{Search for a global optimum.} Consider a form comprising two monomials $m1$ and $m2$. Assume that pre-evaluation is profitable for $m1$ but not for $m2$, and that $m1$ and $m2$ share at least one term (e.g. some basis functions). If pre-evaluation were applied to $m1$, sharing between $m1$ and $m2$ would be lost. We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- results in the highest operation count reduction when considering the whole set of monomials (i.e., the expression as a whole).
\end{enumerate}

Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination. In particular, we define $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, where $\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation, respectively. Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) > 0$. We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}. Having defined $\theta$, we can focus on the transformation algorithm, provided in Algorithm~\ref{algo:gamma}.

\begin{Algo}[Transformation algorithm]
\label{algo:gamma}
\normalfont
The algorithm has three main phases: initialization (step 1); determination of the monomials preserving the memory constraints that should be pre-evaluated (steps 2-4); application of sharing elimination and pre-evaluation (step 5).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree and determine the set of monomials $M$. Let $S$ be the subset of monomials $m$ such that $\theta(m) < 0$. The set of monomials that will \textit{potentially} be pre-evaluated is $P = M \setminus S$. \\ \textit{Note: there are two subtle yet fundamental reasons for not pre-evaluating $m_1 \in P$: 1) the presence of spatial sharing between $m_1$ and $m_2 \in S$, which impacts the search for the global optimum; 2) breaking memory constraints.}
\item Build the set $B$ of all possible bipartitions of $P$.
\item For each $b = (b_S, b_P) \in B$, if the memory required to store the pre-evaluated tables from the monomials in $b_P$ exceeds $T_H$ (see Constraint~\ref{const:TH}), discard $b$; otherwise, add an entry to the dictionary $d$ of potential operation counts, such that $d[b] = \theta_{se}(S \cup b_S) + \theta_{pre}(b_P)$. \\ \textit{Note: $\mathbb{B}$ is in practice very small, since even complex forms usually have only a few monomials. This pass can then be accomplished rapidly as long as the cost of calculating $\theta_{se}$ and $\theta_{pre}$ is negligible. We elaborate on this aspect in Section~\ref{sec:op_count}.}
\item Take $b$ such that $min(d[b])$.
\item Apply pre-evaluation to all monomials in $P \cup b_P$. Apply sharing elimination to the whole expression. \\ \textit{Note: because of reuse of basis functions, pre-evaluation may result in identical tables, which will be mapped to the same temporary. Sharing elimination is therefore transparently applied to the whole expression, including what results from pre-evaluation.}
\end{enumerate}
\end{Algo}

@@

The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the actual transformations and count the number of operations. If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost), the overhead becomes unacceptable when applying pre-evaluation to all possible bipartitions (Algorithm~\ref{algo:pre-evaluation}, due to the symbolic evaluation of $L_i$). We then seek an analytic way of determining $\theta_{pre}$.

The first step consists of estimating the \textit{increase factor}, $\iota$. This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation. To contextualize, consider the example in Figure~\ref{code:increase_factor}. One can think of this as the (simplified) loop nest originating from the integration of a pre-multiplied mass matrix. The sub-expression \texttt{$f_0$*$b_{i0}$+$f_1$*$b_{i1}$+$f_2$*$b_{i2}$} represents the coefficient $f$ over (tabulated) basis functions (array $B$). In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities. By performing product expansion, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor $\iota = 3$.

@@

Despite determining $\iota$ is straightforward in simple scenarios (e.g., when the form has a single coefficient, as we just saw), in general we must account for repeated sub-expressions that, once pre-evaluated, would result in identical tables. This is the case, for example, of a monomial having more than one coefficient in the same function space. One can verify this by taking the example in Figure~\ref{code:increase_factor}, adding a second coefficient $g$ over the same basis functions $b$, and expanding the products. To evaluate $\iota$, we then use combinatorics. We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product (in the example, there is only $f$, so $k=1$); (ii) $n$ is the number of unique basis functions involved in the expansion (in the example, $b_{i0}$, $b_{i1}$, and $b_{i2}$, so $n=3$).

@@

If $\iota < I$, we still need to find the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$. Consider again the mass matrix operator in Figure~\ref{code:increase_factor}. The corresponding monomial is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.  If we instead take the Poisson example in Figure~\ref{code:poisson}, we have that $\rho = 3$ after suitable factorization of basis functions. In general, therefore, $\rho$ depends on both form and discretization employed. To determine this parameter, we re-factorize the expression as per Algorithm~\ref{algo:pre-evaluation} and count the terms amenable to pre-evaluation.

>>>>

\subsection{Transformation algorithm}
We address the two following issues: 
\begin{enumerate}
\item \textit{Coordination of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation could either increase or decrease the operation count with respect to sharing elimination.
\item \textit{Search for a global optimum.} Consider a form comprising two monomials $m_1$ and $m_2$. Assume that pre-evaluation is profitable for $m_1$ but not for $m_2$, and that $m_1$ and $m_2$ share at least one term (e.g. some basis functions). If pre-evaluation were applied to $m_1$, sharing between $m_1$ and $m_2$ would be lost. We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- results in the highest operation count reduction when considering the whole set of monomials (i.e., the expression as a whole).
\end{enumerate}

Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination. In particular, we define $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, where $\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation, respectively. Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) > 0$. We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}. Having defined $\theta$, we can now focus on the transformation algorithm (Algorithm~\ref{algo:gamma}).

\begin{Algo}[Transformation algorithm]
\label{algo:gamma}
\normalfont
The algorithm has three main phases: initialization (step 1); determination of the monomials preserving the memory constraints that should be pre-evaluated (steps 2-4); application of pre-evaluation and sharing elimination (step 5).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree and determine the set of monomials $M$. Let $S$ be the subset of monomials $m$ such that $\theta(m) < 0$. The set of monomials that will \textit{potentially} be pre-evaluated is $P = M \setminus S$. \\ \textit{Note: there are two subtle yet fundamental reasons for not pre-evaluating $m_1 \in P$: 1) the presence of spatial sharing between $m_1$ and $m_2 \in S$, which impacts the search for the global optimum; 2) breaking memory constraints.}
\item Build the set $B$ of all possible bipartitions of $P$.
\item For each $b = (b_S, b_P) \in B$, if the memory required to store the pre-evaluated tables from the monomials in $b_P$ exceeds $T_H$ (see Constraint~\ref{const:TH}), discard $b$; otherwise, add an entry to the dictionary $d$ of potential operation counts, such that $d[b] = \theta_{se}(S \cup b_S) + \theta_{pre}(b_P)$. \\ \textit{Note: $\mathbb{B}$ is in practice very small, since even complex forms usually have only a few monomials. This pass can then be accomplished rapidly as long as the cost of calculating $\theta_{se}$ and $\theta_{pre}$ is negligible. We elaborate on this aspect in Section~\ref{sec:op_count}.}
\item Take $b$ such that $min(d[b])$.
\item Apply pre-evaluation to all monomials in $P \cup b_P$. Apply sharing elimination to the whole expression. \\ \textit{Note: because of reuse of basis functions, pre-evaluation may result in identical tables, which will be mapped to the same temporary. Sharing elimination is therefore transparently applied to the whole expression, including what results from pre-evaluation.}
\end{enumerate}
\end{Algo}

@@

The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the actual transformations and count the number of operations. If, on one hand, this is tolerable for $\theta_{se}$ (Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost), the overhead would be unacceptable if we applied pre-evaluation -- in particular, symbolic execution -- to all bipartitions analyzed by Algorithm~\ref{algo:gamma}. We then seek an analytic way of determining $\theta_{pre}$.

The first step consists of estimating the \textit{increase factor}, $\iota$. This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation. To contextualize, consider the example in Figure~\ref{code:increase_factor}. One can think of this as the (simplified) loop nest originating from the integration of a pre-multiplied mass matrix. The sub-expression \texttt{$f_0$*$b_{i0}$+$f_1$*$b_{i1}$+$f_2$*$b_{i2}$} represents the coefficient $f$ over (tabulated) basis functions (array $B$). In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities. By product expansion, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor $\iota = 3$.

@@

In general, however, determining $\iota$ is not so straightforward since redundant tabulations may result from common sub-expressions. Take the previous example. One can add one coefficient in the same function space as $f$, repeat the expansion, and observe that multiple sub-expressions (e.g., $b_{10}*b_{01}*...$ and $b_{01}*b_{10}*...$) will reduce to identical tables. To evaluate $\iota$, we then use combinatorics. We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product; (ii) $n$ is the number of unique basis functions involved in the expansion. In the original example, we have $n=3$ (for $b_{i0}$, $b_{i1}$, and $b_{i2}$) and $k=1$, which confirms what we intuitively found before, namely $\iota=3$. In the modified example there are two coefficients, so $k=2$, which means $\iota=6$.

@@

If $\iota < I$, we still need to find the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$. Consider again the mass matrix operator in Figure~\ref{code:increase_factor}. The corresponding monomial is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.  If we instead take the Poisson example in Figure~\ref{code:poisson}, we have that $\rho = 3$ after a suitable factorization of basis functions. In general, therefore, $\rho$ depends on both form and discretization employed. To determine this parameter, we re-factorize the expression as per Algorithm~\ref{algo:pre-evaluation} and count the terms amenable to pre-evaluation.
